---
title: "March Madness"
author: "Payton Lyons"
date: "2026-03-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(caret)
library(nnet)
library(randomForest)
library(gbm)
library(pROC)
```

# Data Merging and Cleaning

```{r}
away_neutral <- read.csv("~/Desktop/March Madness 2026/Barttorvik Away-Neutral.csv")
matchups <- read.csv("~/Desktop/March Madness 2026/Tournament Matchups.csv")
```

```{r}
#Organize into one row for each matchup
home <- matchups[seq(1, nrow(matchups), 2), ]
away <- matchups[seq(2, nrow(matchups), 2), ]

games <- data.frame(
  year = home$YEAR,
  round = home$CURRENT.ROUND,
  team1 = home$TEAM,
  seed1 = home$SEED,
  score1 = home$SCORE,
  team2 = away$TEAM,
  seed2 = away$SEED,
  score2 = away$SCORE,
  winner = ifelse(home$SCORE > away$SCORE, home$TEAM, away$TEAM)
)
```

# Logistic Regression

```{r}
model_df <- games %>%
  left_join(away_neutral, by = c("team1" = "TEAM", "year" = "YEAR")) %>%
  rename_with(~ paste0(.x, "_t1"), .cols = -c(year, round, team1, seed1, score1, team2, seed2, score2, winner))

# join barttorvik stats for team2
model_df <- model_df %>%
  left_join(away_neutral, by = c("team2" = "TEAM", "year" = "YEAR")) %>%
  rename_with(~ paste0(.x, "_t2"), .cols = -c(year, round, team1, seed1, score1, team2, seed2, score2, winner, ends_with("_t1")))
```

```{r,results='hide',message=FALSE,warning=FALSE}
model_df <- model_df %>%
  mutate(
    # key efficiency metrics
    badj_o_diff = as.numeric(BADJ.O_t1) - as.numeric(BADJ.O_t2),
    badj_d_diff = as.numeric(BADJ.D_t1) - as.numeric(BADJ.D_t2),
    
    # other
    wab_diff = as.numeric(WAB_t1) - as.numeric(WAB_t2),
    elite_sos_diff = as.numeric(ELITE.SOS_t1) - as.numeric(ELITE.SOS_t2),
    exp_diff = as.numeric(EXP_t1) - as.numeric(EXP_t2),
    talent_diff = as.numeric(TALENT_t1) - as.numeric(TALENT_t2),
    eff_hgt_diff = as.numeric(EFF.HGT_t1) - as.numeric(EFF.HGT_t2),
    seed_diff = seed1 - seed2,
    
    # outcome variable
    outcome = ifelse(winner == team1, 1, 0)
  )

model_df_clean <- na.omit(model_df)
```

```{r}
model_df_train <- model_df_clean %>%
  filter(year < 2025)

model_df_test <- model_df_clean %>%
  filter(year >= 2025)

model_df_train$outcome <- factor(model_df_train$outcome, levels = c(0, 1), labels = c("loss", "win"))
model_df_test$outcome <- factor(model_df_test$outcome, levels = c(0, 1), labels = c("loss", "win"))

model_df_clean$outcome <- factor(model_df_clean$outcome, levels = c(0, 1), labels = c("loss", "win"))

```

```{r}
model <- glm(outcome ~ badj_o_diff+badj_d_diff+wab_diff+elite_sos_diff+
             exp_diff+talent_diff+eff_hgt_diff+seed_diff, data = model_df_train, 
             family = binomial)
summary(model)
```

```{r}
predictions <- predict(model, newdata = model_df_test, type = "response")

predicted_outcome <- ifelse(predictions > 0.45, "win", "loss")

# compare to actual outcomes
actual_outcome <- model_df_test$outcome

# accuracy
mean(predicted_outcome == actual_outcome)

table(predicted = predicted_outcome, actual = actual_outcome)
```

```{r,results = 'hide'}
lmMod <- glm(outcome ~ badj_o_diff+badj_d_diff+wab_diff+elite_sos_diff+
             exp_diff+talent_diff+eff_hgt_diff+seed_diff, data = model_df_train, 
             family = binomial)
selectedMod <- step(lmMod)
```

```{r}
summary(selectedMod)
```

```{r}
predictions_glm <- predict(selectedMod, newdata = model_df_test, type = "response")

predicted_outcome_glm <- ifelse(predictions_glm > 0.5, "win", "loss")

# compare to actual outcomes
actual_outcome <- model_df_test$outcome

# accuracy
mean(predicted_outcome_glm == actual_outcome)

table(predicted = predicted_outcome_glm, actual = actual_outcome)
```

```{r}
data.frame(probability = predictions_glm, outcome = model_df_test$outcome) %>%
  ggplot(aes(x = probability, fill = outcome)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  scale_fill_manual(values = c("loss" = "coral", "win" = "steelblue")) +
  labs(title = "Predicted Win Probability by Actual Outcome",
       x = "Predicted Win Probability", y = "Count", fill = "Actual") +
  geom_vline(xintercept = 0.5,linetype="dashed")+
  theme_bw() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5, color = "grey40"))
```

```{r}
roc_curve <- roc(as.numeric(model_df_test$outcome), predictions)

ggroc(roc_curve, color = "steelblue", linewidth = 1) +
  geom_abline(slope = 1, intercept = 1, linetype = "dashed", color = "grey50") +
  labs(title = "ROC Curve — Logistic Regression",
       subtitle = paste("AUC =", round(auc(roc_curve), 3)),
       x = "Specificity", y = "Sensitivity") +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5, color = "grey40"))
```

```{r}
data.frame(probability = predictions, outcome = as.numeric(model_df_test$outcome) - 1) %>%
  mutate(bin = cut(probability, breaks = seq(0, 1, by = 0.1))) %>%
  group_by(bin) %>%
  summarise(mean_pred = mean(probability),
            mean_actual = mean(outcome),
            n = n()) %>%
  ggplot(aes(x = mean_pred, y = mean_actual)) +
  geom_abline(linetype = "dashed", color = "grey50") +
  geom_point(aes(size = n), color = "steelblue", alpha = 0.8) +
  scale_size_continuous(range = c(2, 8)) +
  labs(title = "Calibration Plot",
       subtitle = "Points on diagonal = perfectly calibrated",
       x = "Mean Predicted Probability",
       y = "Actual Win Rate",
       size = "Games") +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5, color = "grey40"))
```

```{r}
grid <- expand.grid(
  badj_o_diff = seq(min(model_df_test$badj_o_diff), 
                    max(model_df_test$badj_o_diff), length = 100),
  badj_d_diff = seq(min(model_df_test$badj_d_diff), 
                    max(model_df_test$badj_d_diff), length = 100)
)

grid$wab_diff     <- mean(model_df_test$wab_diff)
grid$talent_diff  <- mean(model_df_test$talent_diff)


# logistic regression uses type = "response" and returns probabilities
# so convert using your optimal threshold
grid$pred <- ifelse(predict(selectedMod, newdata = grid, type = "response") > 0.50, 
                    "win", "loss")
grid$pred <- factor(grid$pred, levels = c("loss", "win"))

ggplot() +
  geom_tile(data = grid, aes(x = badj_o_diff, y = badj_d_diff, fill = pred),
            alpha = 0.3) +
  geom_point(data = model_df_test,
             aes(x = badj_o_diff, y = badj_d_diff, color = outcome),
             size = 2) +
  scale_fill_manual(values = c("loss" = "coral", "win" = "lightgreen")) +
  scale_color_manual(values = c("loss" = "darkred", "win" = "forestgreen")) +
  labs(title = "Logistic Regression Decision Boundary",
       subtitle = "Other variables held at mean",
       x = "Offensive Efficiency Diff",
       y = "Defensive Efficiency Diff",
       fill = "Predicted", color = "Actual") +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5, color = "grey40"))
```


# KNN Algorithm

Plan is to test this with both home and neutral/away data.

```{r, results='hide'}
trainControl <- trainControl(method="repeatedcv", number=20, repeats=3)
metric <- "Accuracy"

set.seed(7)
grid <- expand.grid(.k=seq(1,15,by=1))
fit.knn <- train(outcome~badj_o_diff+badj_d_diff+wab_diff+talent_diff, data=model_df_train,
             method="knn", metric=metric,tuneGrid=grid ,trControl=trainControl,
             preProcess = c("center", "scale"))
knn.k1 <- fit.knn$bestTune 
print(fit.knn)
```

```{r}
plot(fit.knn)
```

```{r}
set.seed(7)
prediction <- predict(fit.knn, newdata = model_df_test)
cf <- confusionMatrix(prediction, model_df_test$outcome)
print(cf)
```

```{r}
# create a grid of values spanning your two key variables
grid <- expand.grid(
  badj_o_diff = seq(min(model_df_test$badj_o_diff), 
                    max(model_df_test$badj_o_diff), length = 100),
  badj_d_diff = seq(min(model_df_test$badj_d_diff), 
                    max(model_df_test$badj_d_diff), length = 100)
)

# set other variables to their mean
grid$wab_diff     <- mean(model_df_test$wab_diff)
grid$talent_diff  <- mean(model_df_test$talent_diff)
grid$seed_diff    <- mean(model_df_test$seed_diff)
grid$exp_diff     <- mean(model_df_test$exp_diff)
grid$eff_hgt_diff <- mean(model_df_test$eff_hgt_diff)

grid$pred <- predict(fit.knn, newdata = grid)

ggplot() +
  geom_tile(data = grid, aes(x = badj_o_diff, y = badj_d_diff, fill = pred), 
            alpha = 0.3) +
  geom_point(data = model_df_test, 
             aes(x = badj_o_diff, y = badj_d_diff, color = outcome),
             size = 2) +
  scale_fill_manual(values = c("loss" = "coral", "win" = "lightgreen")) +
  scale_color_manual(values = c("loss" = "darkred", "win" = "forestgreen")) +
  labs(title = "KNN Decision Boundary",
       subtitle = "Other variables held at mean",
       x = "Offensive Efficiency Diff",
       y = "Defensive Efficiency Diff",
       fill = "Predicted", color = "Actual") +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5, color = "grey40"))
```

```{r}
# create a grid of values spanning your two key variables
grid <- expand.grid(
  wab_diff = seq(min(model_df_test$wab_diff), 
                    max(model_df_test$wab_diff), length = 100),
  talent_diff = seq(min(model_df_test$talent_diff), 
                    max(model_df_test$talent_diff), length = 100)
)

# set other variables to their mean
grid$badj_o_diff <- mean(model_df_test$badj_o_diff)
grid$badj_d_diff <- mean(model_df_test$badj_d_diff)

grid$pred <- predict(fit.knn, newdata = grid)

ggplot() +
  geom_tile(data = grid, aes(x = wab_diff, y = talent_diff, fill = pred), 
            alpha = 0.3) +
  geom_point(data = model_df_test, 
             aes(x = wab_diff, y = talent_diff, color = outcome),
             size = 2) +
  scale_fill_manual(values = c("loss" = "coral", "win" = "lightgreen")) +
  scale_color_manual(values = c("loss" = "darkred", "win" = "forestgreen")) +
  labs(title = "KNN Decision Boundary",
       subtitle = "Other variables held at mean",
       x = "Wins Above Bubble Diff",
       y = "Talent Rating Diff",
       fill = "Predicted", color = "Actual") +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5, color = "grey40"))
```

# Neural Network(do more research)

```{r,results='hide'}
nnet_grid <- expand.grid(size = c(1,2,3,4,5),
                         decay = c(0.001, 0.01, 0.1))

fit.nnet <- train(outcome~badj_o_diff+badj_d_diff+wab_diff+talent_diff, 
                  data = model_df_train, method ="nnet",
                  metric = metric, trControl = trainControl,
                  preProcess = c("center", "scale"),
                  tuneGrid = nnet_grid,maxit=500)
```

```{r}
plot(fit.nnet)
```

```{r}
grid <- expand.grid(
  badj_o_diff = seq(min(model_df_test$badj_o_diff), 
                    max(model_df_test$badj_o_diff), length = 100),
  badj_d_diff = seq(min(model_df_test$badj_d_diff), 
                    max(model_df_test$badj_d_diff), length = 100)
)

grid$wab_diff     <- mean(model_df_test$wab_diff)
grid$talent_diff  <- mean(model_df_test$talent_diff)


# logistic regression uses type = "response" and returns probabilities
# so convert using your optimal threshold
grid$pred <- predict(fit.nnet, newdata = grid)

ggplot() +
  geom_tile(data = grid, aes(x = badj_o_diff, y = badj_d_diff, fill = pred),
            alpha = 0.3) +
  geom_point(data = model_df_test,
             aes(x = badj_o_diff, y = badj_d_diff, color = outcome),
             size = 2) +
  scale_fill_manual(values = c("loss" = "coral", "win" = "lightgreen")) +
  scale_color_manual(values = c("loss" = "darkred", "win" = "forestgreen")) +
  labs(title = "Neural Network Decision Boundary",
       subtitle = "Other variables held at mean",
       x = "Offensive Efficiency Diff",
       y = "Defensive Efficiency Diff",
       fill = "Predicted", color = "Actual") +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5, color = "grey40"))
```

# Random Forest
```{r,results='hide'}
fit.rf <- train(outcome~badj_o_diff+badj_d_diff+wab_diff+talent_diff, 
                data = model_df_train, 
             method = "rf", metric = metric,trControl = trainControl,
             preProcess = c("center", "scale"),ntree=500)
```

```{r}
prediction_rf <- predict(fit.rf, newdata = model_df_test)
confusionMatrix(prediction_rf, model_df_test$outcome)
```
```{r}
grid <- expand.grid(
  badj_o_diff = seq(min(model_df_test$badj_o_diff), 
                    max(model_df_test$badj_o_diff), length = 100),
  badj_d_diff = seq(min(model_df_test$badj_d_diff), 
                    max(model_df_test$badj_d_diff), length = 100)
)

grid$wab_diff     <- mean(model_df_test$wab_diff)
grid$talent_diff  <- mean(model_df_test$talent_diff)


# logistic regression uses type = "response" and returns probabilities
# so convert using your optimal threshold
grid$pred <- predict(fit.rf, newdata = grid)

ggplot() +
  geom_tile(data = grid, aes(x = badj_o_diff, y = badj_d_diff, fill = pred),
            alpha = 0.3) +
  geom_point(data = model_df_test,
             aes(x = badj_o_diff, y = badj_d_diff, color = outcome),
             size = 2) +
  scale_fill_manual(values = c("loss" = "coral", "win" = "lightgreen")) +
  scale_color_manual(values = c("loss" = "darkred", "win" = "forestgreen")) +
  labs(title = "Random Forest Decision Boundary",
       subtitle = "Other variables held at mean",
       x = "Offensive Efficiency Diff",
       y = "Defensive Efficiency Diff",
       fill = "Predicted", color = "Actual") +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5, color = "grey40"))
```

# Gradient Boosting
```{r,results='hide'}
fit.gbm <- train(outcome ~badj_o_diff+badj_d_diff+wab_diff+talent_diff, 
             data = model_df_train, method = "gbm", metric = metric, 
             trControl = trainControl, preProcess = c("center", "scale"))
```

```{r}
prediction_gbm <- predict(fit.gbm, newdata = model_df_test)
confusionMatrix(prediction_gbm, model_df_test$outcome)
```

```{r}
grid <- expand.grid(
  badj_o_diff = seq(min(model_df_test$badj_o_diff), 
                    max(model_df_test$badj_o_diff), length = 100),
  badj_d_diff = seq(min(model_df_test$badj_d_diff), 
                    max(model_df_test$badj_d_diff), length = 100)
)

grid$wab_diff     <- mean(model_df_test$wab_diff)
grid$talent_diff  <- mean(model_df_test$talent_diff)


# logistic regression uses type = "response" and returns probabilities
# so convert using your optimal threshold
grid$pred <- predict(fit.gbm, newdata = grid)

ggplot() +
  geom_tile(data = grid, aes(x = badj_o_diff, y = badj_d_diff, fill = pred),
            alpha = 0.3) +
  geom_point(data = model_df_test,
             aes(x = badj_o_diff, y = badj_d_diff, color = outcome),
             size = 2) +
  scale_fill_manual(values = c("loss" = "coral", "win" = "lightgreen")) +
  scale_color_manual(values = c("loss" = "darkred", "win" = "forestgreen")) +
  labs(title = "Gradient Boosting Method Decision Boundary",
       subtitle = "Other variables held at mean",
       x = "Offensive Efficiency Diff",
       y = "Defensive Efficiency Diff",
       fill = "Predicted", color = "Actual") +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5, color = "grey40"))
```
# Summary and Model Selection

Selected the reduced logistic regression to create tournament bracket.

# 2026 Bracket Predictions

```{r}
bracket_model <- glm(outcome ~ badj_o_diff+badj_d_diff+wab_diff+talent_diff, 
                     data = model_df, family = binomial)
summary(bracket_model)
```

```{r}
df_26 <- away_neutral %>%
  filter(YEAR == 2026) %>%
  select(c("TEAM","BADJ.O","BADJ.D","WAB","TALENT"))
```
## Logistic Regression Predictions

```{r}
predict_game_lm <- function(team1, team2) {
  teams <- c(team1, team2)
  game <- df_26 %>% filter(TEAM %in% teams)
  game_df <- game[game$TEAM == team1, 2:5] - game[game$TEAM == team2, 2:5]
  colnames(game_df) <- c("badj_o_diff", "badj_d_diff", "wab_diff", "talent_diff")
  prob <- predict(bracket_model, newdata = game_df, type = "response")
  winner <- ifelse(prob > 0.5, team1, team2)
  win_prob <- round(ifelse(prob > 0.5, prob, 1 - prob) * 100, 1)
  cat(team1, "vs", team2, "-->", winner, "(", win_prob, "%)\n")
}

# enter all first round matchups
matchups <- list(
  c("Duke", "Siena"),
  c("Ohio St.", "TCU"),
  c("St. John's", "Northern Iowa"),
  c("Kansas", "Cal Baptist"),
  c("Louisville", "South Florida"),
  c("Michigan St.", "North Dakota St."),
  c("UCLA", "UCF"),
  c("Connecticut", "Furman"),
  c("Florida", "Prairie View A&M"),
  c("Clemson", "Iowa"),
  c("Vanderbilt", "McNeese St."),
  c("Nebraska", "Troy"),
  c("North Carolina", "VCU"),
  c("Illinois", "Penn"),
  c("Saint Mary's", "Texas A&M"),
  c("Houston", "Idaho"),
  c("Arizona", "LIU Brooklyn"),
  c("Villanova", "Utah St."),
  c("Wisconsin", "High Point"),
  c("Arkansas", "Hawaii"),
  c("BYU", "Texas"),
  c("Gonzaga", "Kennesaw St."),
  c("Miami FL", "Missouri"),
  c("Purdue", "Queens"),
  c("Michigan", "Howard"),
  c("Georgia", "Saint Louis"),
  c("Texas Tech", "Akron"),
  c("Alabama", "Hofstra"),
  c("Tennessee", "Miami OH"),
  c("Virginia", "Wright St."),
  c("Kentucky", "Santa Clara"),
  c("Iowa St.", "Tennessee St.")
  
)
```
#### Round 1
```{r}
# run all predictions at once
cat("=== First Round Predictions ===\n\n")
invisible(lapply(matchups, function(m) predict_game_lm(m[1], m[2])))
```
#### Round 2

```{r}
matchups <- list(
  c("Duke", "Ohio St."),
  c("Kansas", "St. John's"),
  c("Michigan St.", "Louisville"),
  c("Connecticut", "UCLA"),
  c("Florida", "Clemson"),
  c("Nebraska", "Vanderbilt"),
  c("Illinois", "VCU"),
  c("Houston", "Texas A&M"),
  c("Arizona", "Villanova"),
  c("Arkansas", "Wisconsin"),
  c("Gonzaga", "BYU"),
  c("Purdue", "Miami FL"),
  c("Michigan", "Georgia"),
  c("Alabama", "Texas Tech"),
  c("Virginia", "Tennessee"),
  c("Iowa St.", "Kentucky")
)

cat("=== Second Round Predictions ===\n\n")
invisible(lapply(matchups, function(m) predict_game_lm(m[1], m[2])))
```

#### Sweet 16

```{r}
matchups <- list(
  c("Duke", "St. John's"),
  c("Michigan St.", "Connecticut"),
  c("Florida", "Vanderbilt"),
  c("Houston", "Illinois"),
  c("Arizona", "Arkansas"),
  c("Purdue", "Gonzaga"),
  c("Michigan", "Alabama"),
  c("Iowa St.", "Virginia")
)

cat("=== Sweet 16 Predictions ===\n\n")
invisible(lapply(matchups, function(m) predict_game_lm(m[1], m[2])))
```

#### Elite 8

```{r}
matchups <- list(
  c("Duke", "Connecticut"),
  c("Florida", "Houston"),
  c("Arizona", "Purdue"),
  c("Michigan", "Virginia")
)

cat("=== Sweet 16 Predictions ===\n\n")
invisible(lapply(matchups, function(m) predict_game_lm(m[1], m[2])))
```

#### Semi-Finals

```{r}
matchups <- list(
  c("Duke", "Houston"),
  c("Houston", "Duke"),
  c("Arizona", "Michigan"),
  c("Michigan", "Arizona")
)

cat("=== Semifinal Predictions ===\n\n")
invisible(lapply(matchups, function(m) predict_game_lm(m[1], m[2])))
```
#### Finals

```{r}
matchups <- list(
  c("Duke", "Michigan"),
  c("Michigan", "Duke")
)

cat("=== Final Predictions ===\n\n")
invisible(lapply(matchups, function(m) predict_game_lm(m[1], m[2])))
```

## Random Forest Predictions (PnzvWL)
```{r, results = 'hide'}
modelrf <- train(outcome~badj_o_diff+badj_d_diff+wab_diff+talent_diff, 
                data = model_df_clean, 
             method = "rf", metric = metric,trControl = trainControl,
             preProcess = c("center", "scale"),ntree=500)
                  
predict_game_rf <- function(team1, team2) {
  teams <- c(team1, team2)
  game <- df_26 %>% filter(TEAM %in% teams)
  game_df <- game[game$TEAM == team1, 2:5] - game[game$TEAM == team2, 2:5]
  colnames(game_df) <- c("badj_o_diff", "badj_d_diff", "wab_diff", "talent_diff")
  prob <- predict(modelrf, newdata = game_df, type = "prob")[,2]
  winner <- ifelse(prob > 0.5, team1, team2)
  win_prob <- round(ifelse(prob > 0.5, prob, 1 - prob) * 100, 1)
  cat(team1, "vs", team2, "-->", winner, "(", win_prob, "%)\n")
}
```
#### Round 1
```{r}
matchups <- list(
  c("Duke", "Siena"),
  c("Ohio St.", "TCU"),
  c("St. John's", "Northern Iowa"),
  c("Kansas", "Cal Baptist"),
  c("Louisville", "South Florida"),
  c("Michigan St.", "North Dakota St."),
  c("UCLA", "UCF"),
  c("Connecticut", "Furman"),
  c("Florida", "Prairie View A&M"),
  c("Clemson", "Iowa"),
  c("Vanderbilt", "McNeese St."),
  c("Nebraska", "Troy"),
  c("North Carolina", "VCU"),
  c("Illinois", "Penn"),
  c("Saint Mary's", "Texas A&M"),
  c("Houston", "Idaho"),
  c("Arizona", "LIU Brooklyn"),
  c("Villanova", "Utah St."),
  c("Wisconsin", "High Point"),
  c("Arkansas", "Hawaii"),
  c("BYU", "Texas"),
  c("Gonzaga", "Kennesaw St."),
  c("Miami FL", "Missouri"),
  c("Purdue", "Queens"),
  c("Michigan", "Howard"),
  c("Georgia", "Saint Louis"),
  c("Texas Tech", "Akron"),
  c("Alabama", "Hofstra"),
  c("Tennessee", "Miami OH"),
  c("Virginia", "Wright St."),
  c("Kentucky", "Santa Clara"),
  c("Iowa St.", "Tennessee St.")
  
)

cat("=== First Round Predictions ===\n\n")
invisible(lapply(matchups, function(m) predict_game_rf(m[1], m[2])))
```
#### Round 2
```{r}
matchups <- list(
  c("Duke", "Ohio St."),
  c("Kansas", "St. John's"),
  c("Michigan St.", "Louisville"),
  c("Connecticut", "UCLA"),
  c("Florida", "Clemson"),
  c("Nebraska", "Vanderbilt"),
  c("Illinois", "VCU"),
  c("Houston", "Saint Mary's"),
  c("Arizona", "Villanova"),
  c("Arkansas", "Wisconsin"),
  c("Gonzaga", "BYU"),
  c("Purdue", "Miami FL"),
  c("Michigan", "Georgia"),
  c("Alabama", "Texas Tech"),
  c("Virginia", "Tennessee"),
  c("Iowa St.", "Kentucky")
)

cat("=== Second Round Predictions ===\n\n")
invisible(lapply(matchups, function(m) predict_game_rf(m[1], m[2])))
```
#### Sweet 16

```{r}
matchups <- list(
  c("Duke", "St. John's"),
  c("Connecticut", "Michigan"),
  c("Florida", "Vanderbilt"),
  c("Houston", "Illinois"),
  c("Arizona", "Arkansas"),
  c("Purdue", "Gonzaga"),
  c("Michigan", "Alabama"),
  c("Iowa St.", "Tennessee")
)

cat("=== Sweet 16 Predictions ===\n\n")
invisible(lapply(matchups, function(m) predict_game_rf(m[1], m[2])))
```

#### Elite 8

```{r}
matchups <- list(
  c("Duke", "Connecticut"),
  c("Vanderbilt", "Illinois"),
  c("Arizona", "Purdue"),
  c("Michigan", "Virginia")
)

cat("=== Sweet 16 Predictions ===\n\n")
invisible(lapply(matchups, function(m) predict_game_rf(m[1], m[2])))
```

#### Semi-Finals
Predict each game twice, one with each team being the higher seed then average the predicted probabilities.

```{r}
matchups <- list(
  c("Duke", "Illinois"),
  c("Illinois", "Duke"),
  c("Arizona", "Michigan"),
  c("Michigan", "Arizona")
)

cat("=== Semifinal Predictions ===\n\n")
invisible(lapply(matchups, function(m) predict_game_rf(m[1], m[2])))
```

#### Finals

```{r}
matchups <- list(
  c("Duke", "Michigan"),
  c("Michigan", "Duke")
)

cat("=== Final Predictions ===\n\n")
invisible(lapply(matchups, function(m) predict_game_rf(m[1], m[2])))
```

## Gradient Boosting Method Predictions (H48Hsd)
```{r, results = 'hide'}
modelgbm <- train(outcome ~badj_o_diff+badj_d_diff+wab_diff+talent_diff, 
             data = model_df_clean, method = "gbm", metric = metric, 
             trControl = trainControl, preProcess = c("center", "scale"))
                  
predict_game_gbm <- function(team1, team2) {
  teams <- c(team1, team2)
  game <- df_26 %>% filter(TEAM %in% teams)
  game_df <- game[game$TEAM == team1, 2:5] - game[game$TEAM == team2, 2:5]
  colnames(game_df) <- c("badj_o_diff", "badj_d_diff", "wab_diff", "talent_diff")
  prob <- predict(modelgbm, newdata = game_df, type = "prob")[,2]
  winner <- ifelse(prob > 0.5, team1, team2)
  win_prob <- round(ifelse(prob > 0.5, prob, 1 - prob) * 100, 1)
  cat(team1, "vs", team2, "-->", winner, "(", win_prob, "%)\n")
}
```
#### Round 1
```{r}
matchups <- list(
  c("Duke", "Siena"),
  c("Ohio St.", "TCU"),
  c("St. John's", "Northern Iowa"),
  c("Kansas", "Cal Baptist"),
  c("Louisville", "South Florida"),
  c("Michigan St.", "North Dakota St."),
  c("UCLA", "UCF"),
  c("Connecticut", "Furman"),
  c("Florida", "Prairie View A&M"),
  c("Clemson", "Iowa"),
  c("Vanderbilt", "McNeese St."),
  c("Nebraska", "Troy"),
  c("North Carolina", "VCU"),
  c("Illinois", "Penn"),
  c("Saint Mary's", "Texas A&M"),
  c("Houston", "Idaho"),
  c("Arizona", "LIU Brooklyn"),
  c("Villanova", "Utah St."),
  c("Wisconsin", "High Point"),
  c("Arkansas", "Hawaii"),
  c("BYU", "Texas"),
  c("Gonzaga", "Kennesaw St."),
  c("Miami FL", "Missouri"),
  c("Purdue", "Queens"),
  c("Michigan", "Howard"),
  c("Georgia", "Saint Louis"),
  c("Texas Tech", "Akron"),
  c("Alabama", "Hofstra"),
  c("Tennessee", "Miami OH"),
  c("Virginia", "Wright St."),
  c("Kentucky", "Santa Clara"),
  c("Iowa St.", "Tennessee St.")
  
)

cat("=== First Round Predictions ===\n\n")
invisible(lapply(matchups, function(m) predict_game_gbm(m[1], m[2])))
```

#### Round 2
```{r}
matchups <- list(
  c("Duke", "TCU"),
  c("Kansas", "St. John's"),
  c("Michigan St.", "Louisville"),
  c("Connecticut", "UCLA"),
  c("Florida", "Clemson"),
  c("Nebraska", "Vanderbilt"),
  c("Illinois", "VCU"),
  c("Houston", "Texas A&M"),
  c("Arizona", "Villanova"),
  c("Arkansas", "Wisconsin"),
  c("Gonzaga", "BYU"),
  c("Purdue", "Miami FL"),
  c("Michigan", "Georgia"),
  c("Alabama", "Texas Tech"),
  c("Virginia", "Tennessee"),
  c("Iowa St.", "Kentucky")
)

cat("=== Second Round Predictions ===\n\n")
invisible(lapply(matchups, function(m) predict_game_gbm(m[1], m[2])))
```

#### Sweet 16

```{r}
matchups <- list(
  c("Duke", "St. John's"),
  c("Connecticut", "Michigan St."),
  c("Florida", "Vanderbilt"),
  c("Houston", "Illinois"),
  c("Arizona", "Arkansas"),
  c("Purdue", "Gonzaga"),
  c("Michigan", "Alabama"),
  c("Iowa St.", "Tennessee")
)

cat("=== Sweet 16 Predictions ===\n\n")
invisible(lapply(matchups, function(m) predict_game_gbm(m[1], m[2])))
```

#### Elite 8

```{r}
matchups <- list(
  c("Duke", "Connecticut"),
  c("Florida", "Houston"),
  c("Arizona", "Purdue"),
  c("Michigan", "Virginia")
)

cat("=== Sweet 16 Predictions ===\n\n")
invisible(lapply(matchups, function(m) predict_game_gbm(m[1], m[2])))
```

#### Semi-Finals

```{r}
matchups <- list(
  c("Duke", "Houston"),
  c("Houston", "Duke"),
  c("Arizona", "Michigan"),
  c("Michigan", "Arizona")
)

cat("=== Semifinal Predictions ===\n\n")
invisible(lapply(matchups, function(m) predict_game_gbm(m[1], m[2])))
```

#### Finals

```{r}
matchups <- list(
  c("Duke", "Michigan"),
  c("Michigan", "Duke")
)

cat("=== Final Predictions ===\n\n")
invisible(lapply(matchups, function(m) predict_game_gbm(m[1], m[2])))
```
